\\LARGE Decision-Optimal Probabilistic Forecast Reconciliation for Expected Shortfall in Hierarchical Time Series

### 2025-08-27 phd_application/phd_propossal.Rmd
update bkgroud and motiv

These tail risks are, however, not managed in isolation. Consider a hedge fund's investment portfolio returns, which consists returns of different asset classes, and further contains returns of various individual securities. This is an example of a hierarchical structure, with multiple *aggregation levels*. Risk managers need to assess and forecast the risk quantities at several levels, from the overall portfolio, down to individual securities. Ignoring this naturally-formed hierarchical structure can lead to misaligned forecasts, for example, the forecasted returns of individual securities are low while the overall portfolio return is high. This results in misallocation of capital and suboptimal risk management strategies.

Forecasting such hierarchical structure also arises in many other decision-making contexts, from supply chains [@Seaman2022-bb; @Angam2025-od] and energy planning [@Di-Modica2021-ad], to macroeconomic [@El-Gemayel2022-hx; @Li2019-vv] and tourism analysis [@Athanasopoulos2009-lp]. Stakeholders in these settings need forecasts at several aggregation levels to allocate resources and manage risk. These hierarchical structures naturally form *aggregation constraints* (child levels must sum up to parents). Taken together, when we have a collection of time series organised in a hierarchy subject to aggregation constraints, we refer to this as *hierarchical time series*.

In practice, when forecasts are produced for all series (often called *base forecasts*), they typically violate the aggregation constraints observed in the data; such forecasts are *incoherent* (e.g., security-level numbers don’t sum to the overall trading book). This can undermine downstream decisions that require internal consistency. Forecast reconciliation, a post-processing step, addresses this by adjusting the base forecasts so the final adjusted set of forecasts is *coherent* with the aggregation structure. Approaches such as OLS [@Hyndman2011-jv], WLS [@Hyndman2016-ic], MinT [@Wickramasuriya2019-xq], and more are developed and commonly used in practice. An additional incentive for the rapid adoption of reconciliation methods is that they improve forecast accuracy [@Athanasopoulos2024-as]. For instance, the International Monetary Fund reported improvements in their liquidity forecasts, generating economic benefits via more efficient monetary policy operations [@El-Gemayel2022-hx].

\color{red}
Importantly, ES behaves well under aggregation (satisfying subadditivity\footnote{\textbf{Subadditivity.} A risk measure $\rho$ on loss variables is subadditive if $\rho(X+Y)\le \rho(X)+\rho(Y)$ for any losses $X,Y$. It encodes diversification: the risk of a combined portfolio should not exceed the sum of stand-alone risks. Expected Shortfall satisfies subadditivity for all $\alpha\in(0,1)$, whereas VaR can violate it. Thus, the portfolio ES cannot exceed the sums of desk ESs.}), making it a natural target when risks are rolled up through a hierarchy. This context motivates methods that deliver coherent, multi-level forecasts whose uncertainty aligns with the tail-risk metrics managers actually use.

@@

Modern risk management and operations need more than a typical mean or median value--they need quantities from the full predictive distribution that capture uncertainty and tail behaviour, because many decisions are tail-sensitive (e.g., extreme demand in retail settings or energy generation shortfalls). While VaR is conceptually straightforward, it ignores the severity of losses beyond the VaR threshold and is not a coherent risk measure, since it does not satisfy the subadditivity property [@Yamai2005-me]. VaR can mislead risk assessments across different levels of the hierarchy, and may underestimate the true risk exposure under market stress. ES, on the other hand, is a coherent metric that captures the average loss in the tail beyond the VaR threshold. This makes ES a more reliable metric for assessing and managing "tail risks", and it has been adopted as the preferred regulatory metric since the Basel III framework [@Basel-iii]. If ES is the metric that risk managers actually use, reconciliation methods should guarantee the coherence and optimality with respect to ES.

Most reconciliation methodology has been developed for point forecasts and are optimised for point scoring functions such as mean squared error (MSE). Generalising these methods to reconcile probabilistic forecasts does not guarantee optimality for tail-risk measures such as VaR or ES. Although @Panagiotelis2023-fs introduced an score optimisation for reconciliation weights using proper scoring rules (energy score and variogram score) for full multivariate predictive distributions, they do not directly address the tail functionals. This leaves the question: *"How can we construct probabilistic forecast reconciliation that are explicitly optimised for tail-risk measures such as ES?"*

@@

As a result, the objective of this project is to bring these theoretical advances into the reconciliation framework and develop methods that produce coherent probabilistic forecasts optimised for ES. Our goal is a reliable, consistent, and decision-aligned risk-forecasting framework that helps national and international organisations improve ES backtesting, stabilise capital, and allocate resources more effectively across hierarchical levels.

>>>>

These tail risks are, however, not managed in isolation. Consider a hedge fund's investment portfolio returns, which compose of returns of different asset classes, and each asset class further contain returns of various individual securities. This is an example of a hierarchical structure, with multiple *aggregation levels*. Risk managers need to assess and forecast the returns or risk quantities at several levels, from the overall portfolio, down to individual securities. Ignoring this naturally-formed hierarchical structure can lead to misaligned forecasts, for example, the predicted returns of individual securities are low while the overall portfolio returns are high. This results in misallocation of capital and suboptimal risk management strategies.

Forecasting such hierarchical structure also arises in many other decision-making contexts, from supply chains [@Seaman2022-bb; @Angam2025-od] and energy planning [@Di-Modica2021-ad], to macroeconomic [@El-Gemayel2022-hx; @Li2019-vv] and tourism analysis [@Athanasopoulos2009-lp]. Stakeholders in these settings need forecasts at several aggregation levels to allocate resources and manage risk. These hierarchical structures naturally form *aggregation constraints* (child levels must sum up to parents). Taken together, when we have a collection of time series organised in a hierarchy subject to the aggregation constraints, we refer to this as *hierarchical time series*.

In practice, when forecasts are produced for all series (often called *base forecasts*), they typically violate the aggregation constraints observed in the data; such forecasts are *incoherent*. This can undermine downstream decisions that require internal consistency. Forecast reconciliation, a post-processing step, addresses this by adjusting the base forecasts so the final adjusted set of forecasts is *coherent* with the aggregation structure. Approaches such as OLS [@Hyndman2011-jv], WLS [@Hyndman2016-ic], MinT [@Wickramasuriya2019-xq], and more are developed and commonly used in practice. An additional incentive for the rapid adoption of reconciliation methods is that they improve forecast accuracy [@Athanasopoulos2024-as]. For instance, the International Monetary Fund reported improvements in their liquidity forecasts, generating economic benefits via more efficient monetary policy operations [@El-Gemayel2022-hx].

\color{red}
Importantly, ES behaves well under aggregation (satisfying subadditivity), making it a natural target when risks are rolled up through a hierarchy. This context motivates methods that deliver coherent, multi-level forecasts whose uncertainty aligns with the tail-risk metrics managers actually use.

@@

Modern risk management and operations need more than a typical mean or median value--they need quantities from the full predictive distribution that capture uncertainty and tail behaviour, because many decisions are tail-sensitive (e.g., extreme demand in retail settings or energy generation shortfalls). While VaR is conceptually straightforward, it ignores the severity of losses beyond the VaR threshold and does not satisfy the subadditivity property\footnote{\textbf{Subadditivity.} A risk measure $\rho$ on loss variables is subadditive if $\rho(X+Y)\le \rho(X)+\rho(Y)$ for any losses $X,Y$. Expected Shortfall satisfies subadditivity for all $\alpha\in(0,1)$, whereas VaR can violate it. Thus, the portfolio ES cannot exceed the sums of individual security ESs.} [@Yamai2005-me]. Violating subadditivity is a critical problem, because it means that the risk measure of a combined portfolio can exceed the sum of stand-alone risks, which contradicts the principle of diversification. VaR therefore can mislead risk assessments across different levels of the hierarchy, and may underestimate the true risk exposure under market stress. ES, on the other hand, is a consistent metric that captures the average loss in the tail beyond the VaR threshold. This makes ES a more reliable measure for assessing and managing "tail risks", and it has been adopted as the preferred regulatory metric since the Basel III framework [@Basel-iii]. If ES is the metric that risk managers actually use, reconciliation methods should guarantee the coherence and optimality with respect to ES.

Most reconciliation methodology has been developed for point forecasts and are optimised for point scoring functions such as mean squared error (MSE). Generalising these methods to reconcile probabilistic forecasts does not guarantee optimality for tail-risk measures such as VaR or ES. Although @Panagiotelis2023-fs introduced score optimisation for reconciliation weights using proper scoring rules (energy score and variogram score) for full multivariate predictive distributions, they do not directly address the tail functionals. This leaves the question: *"How can we construct probabilistic forecast reconciliation that are explicitly optimised for tail-risk measures such as ES?"*

<!-- \color{red}
**AP: I have thought a bit about how we discuss the point v probabilistic issue. Ultimately, a VaR or ES forecast is a point forecast. However, a) any point forecast can be derived from a distributional forecast and b) scoring rules give us a way to obtain the optimal VaR or ES if we know the underlying probabilistic distribution. These two facts taken together mean we can naturally adapt the score optimisation framework of Panagiotelis et al (2023) to obtain optimal ES forecasts while also guaranteeing that these forecast come from an underlying probabilitistic forecast that is coherent.**
\color{black} -->

@@

As a result, the objective of my thesis is to bring these theoretical advances into the reconciliation framework and develop methods that produce coherent probabilistic forecasts optimised for ES. The goal is a reliable, consistent, and decision-aligned risk-forecasting framework that helps national and international organisations improve ES backtesting, stabilise capital, and allocate resources more effectively across hierarchical levels.
